---
title: "Modeling cell type colocalization in multiplex imaging data using functional regression and wild bootstrap"
author: "Carly Middleton, Qi Zheng, Jun Yan, Susmita Datta, Maiying Kong"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the R package `funboot`.  In the following, we demonstrate how to 1.) use in-package functions to preprocess single-cell spatial data, 2.) compute spatial summary functions and colocalization curves for pairs of cell types, and 3.) quantify colocalization as a function of patient-level and spatial covariates, using the concurrent functional regression framework and our wild bootstrap-based inference procedure.  

We illustrate this workflow by reproducing our manuscript's real data analyses of the publicly available [Jackson-Fischer breast cancer](https://bodenmillergroup.github.io/imcdatasets/reference/JacksonFischer_2020_BreastCancer.html) and [Hoch-Schulz melanoma](https://bodenmillergroup.github.io/imcdatasets/reference/HochSchulz_2022_Melanoma.html) imaging mass cytometry datasets.

Although a minimum of `B1 = 50` outer and `B2 = 50` inner bootstrap samples are generally recommended in practice, this vignette uses `B1 = B2 = 10` and `n_perm=` to reduce computation time.  As a result, the vignette takes about 4 minutes to build on a standard laptop computer with 8 cores, a 2.3 GHz processor, and 16 GB of RAM.  



## Installation


The package's dependencies are below. 

```{r, error=FALSE, warning=FALSE, message=FALSE}
library(mgcv)
library(refund)
library(spatstat.geom)
library(spatstat.explore)
library(ggplot2)
```

`funboot` can be installed and loaded as follows:

```{r, eval=FALSE, error=FALSE, warning=FALSE, message=FALSE}
devtools::install_github('carlyemiddleton/funboot', build_vignettes = TRUE, force=TRUE)
```

```{r, error=FALSE, warning=FALSE, message=FALSE}
library(funboot)
```

Though not a package dependency, we also use `patchwork` within the vignette to assist with plotting:

```{r, error=FALSE, warning=FALSE, message=FALSE}
library(patchwork)
```

## Breast Cancer Data

The 100-image subset of the Jackson-Fischer data contains 1 image per subject.  It is stored in-package and can be loaded as follows:  

```{r }
data("breastcancer_data", package='funboot')
```

We focus on cell metaclusters 3 (T cells), 7 (endothelial cells), 9 (small circular stromal cells), and 16 (proliferative epithelial cells).  Their spatial distributions can be examined using `funboot::plot_images()`:

```{r fig1, fig.cap="**Figure 1.** Spatial distributions of selected cell types in representative breast cancer images. Colors represent cell phenotype.", fig.width = 7, fig.height = 12, warning=FALSE, message=FALSE}
##Choose colors for each cell type
palette <- c('Other' = "gray", 
             '7' = 'gold', 
             '3' = "darkgreen", 
             '9' = 'skyblue', 
             '16'='cyan4')
##Plot the images featured in the manuscript
plot_list <- funboot::plot_images(data = breastcancer_data,
                                  images_to_plot = c(13,181,246,362,219,244,246,254), 
                                  cell_types_to_plot = c(7,3,9,16),
                                  palette = palette)
##Each entry of plot_list corresponds to one of images_to_plot
(plot_list[[1]] + plot_list[[2]])/
(plot_list[[3]] + plot_list[[4]])/
(plot_list[[5]] + plot_list[[6]])/
(plot_list[[7]] + plot_list[[8]])

```

Cell types 7 (endothelial cells) and 3 (T cells) appear to be colocalized.  We can examine their colocalization curve by fitting the model

$$
Y_{i}(r) = \beta_0(r) + \epsilon_i(r), \qquad i = 1, \ldots, n,
$$
where $Y_{i}(r)$ denotes the observed colocalization curve for T cells around endothelial cells within the image belonging to subject $i$, and performing inference on the intercept $\beta_0(r)$.  The first step is to discard images with sparse cell counts and then calculate the observed and expected spatial summary function.  Because some of these images contain holes, we apply the permutation adjustment during the calculation of the expected spatial summary function.  The process can be carried out using `funboot::preprocess_data()`.  


```{r results='hide', message=FALSE, warning=FALSE, error=FALSE}
set.seed(12345)
sumfun_data_interaction <- funboot::preprocess_data(data=breastcancer_data, 
                                                    from_cell=7, 
                                                    to_cell=3,
                                                    qc_cellcount_cutoff=20, 
                                                    n_perm=5, 
                                                    perm_yn=TRUE,
                                                    r_max=500, 
                                                    inc=1, 
                                                    image_dims=c(0,1000,0,1000),
                                                    summary_function='L')

```

we can calculate and plot a confidence band for the sum of both intercepts using `funboot::wildBS_CB()` and `funboot::plot_wildBS_CB()`.

```{r fig.width = 7, fig.height = 3, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
CB_interaction <- funboot::wildBS_CB(formula=outcome ~ 1,
                        data=sumfun_data_interaction,
                        spatial_covars = NULL,
                        target = c("Intercept"),
                        form   = c("coef"),
                        covar_list = list(),
                        func = function(Intercept){
                          Intercept
                        },
                        B1=10,
                        B2=10,
                        alpha=.05,
                        re=NULL,
                        nthreads = 1,
                        id='patient_id')
```

```{r fig2, fig.cap="**Figure 2.** Spatial distributions of selected cell types in representative breast cancer images.", fig.width = 7, fig.height = 3, warning=FALSE, message=FALSE, results='hide'}
##Base plot
p <- funboot::plot_wildBS_CB(CB_interaction)
##Layer for the image-specific curves
bg_df <- data.frame(
  r = sumfun_data_interaction$r,
  y = sumfun_data_interaction$outcome,
  curve = sumfun_data_interaction$image_number
)
bg_layer <- geom_line(
  data = bg_df,
  aes(x = r, y = y, group = curve, color = "Observed Y(r)"),
  inherit.aes = FALSE,
  linewidth = 0.3
)
##Combine both layers and add formatting
p$layers <- c(list(bg_layer), p$layers)
p + ggtitle("Colocalization Curve") + 
  labs(
    x = "r",
    y = "Y(r)"
  ) + 
    scale_color_manual(
    values = c(
      "Observed Y(r)" = "gray80",
      "Confidence Band" = "#00BA38",
      "Estimated Target Curve" = "#F8766D"
    ),
    labels = c(
      "Observed Yj(r)" = "Observed Y(r)",
      "Estimated Target Curve" = "Estimated E[Y(r)]",
      "Confidence Band" = "95% Confidence Band"
    ),
    breaks = c("Confidence Band", "Estimated Target Curve", "Observed Y(r)")
  )

```

Because the confidence band crosses zero, the conclusion would be that a significant amount of interaction is present. (However, many more bootstrap samples are needed for accurate results.)  The above can also be repeated for the avoidance example.

Next, we incorporate covariates into the model:

```{r message=FALSE, warning=FALSE, error=FALSE, results='hide'}
sumfun_data_covariates <- preprocess_data(data=breastcancer_data,
                                           from_cell=7,
                                           to_cell=3,
                                           qc_cellcount_cutoff=20,
                                           n_perm=5,
                                           perm_yn=T,
                                           r_max=500,
                                           inc=1,
                                           image_dims=c(0,1000,0,1000),
                                           summary_function='L',
                                           verbose=TRUE)
#combine cell types 3,4,5,6 into 1 category
#breastcancer_data$cell_type <- ifelse(breastcancer_data$cell_type %in% c(3:6), 3,
#                                                       breastcancer_data$cell_type)
spatialcov_data <- preprocess_data(data=breastcancer_data,
                                   from_cell=7,
                                   to_cell=7,
                                   qc_cellcount_cutoff=20,
                                   n_perm=5,
                                   perm_yn=T,
                                   r_max=500,
                                   inc=1,
                                   image_dims=c(0,1000,0,1000),
                                   summary_function='L',
                                   verbose=TRUE)
names(spatialcov_data)[names(spatialcov_data)=='L_obs'] <- 'spatial_cov'
spatialcov_data$L_expect <- spatialcov_data$L_pmean <- spatialcov_data$outcome <- NULL
sumfun_data_full_covariates <- merge(sumfun_data_covariates,
                                     spatialcov_data,
                                     by=c('image_number','r','patient_id','patient_age','tumor_grade'),
                                     all=T)
df <- data.frame(image_number = rep(unique(breastcancer_data$image_number), each=501),
                 r = rep(0:500, length(unique(breastcancer_data$image_number)))
                 )
sumfun_data_full_covariates <- merge(df,
                                     sumfun_data_full_covariates,
                                     by=c('image_number','r'),
                                     all=T)
sumfun_data_full_covariates <- sumfun_data_full_covariates[!is.na(sumfun_data_full_covariates$outcome),] #Drop obs. with missing outcomes
sumfun_data_full_covariates$spatial_cov <- ifelse(is.na(sumfun_data_full_covariates$spatial_cov), 0,
                                                  sumfun_data_full_covariates$spatial_cov) #set L^{3,16}(r) = 0 when there are no type 16 cells
sumfun_data_full_covariates$grade2 <- ifelse(sumfun_data_full_covariates$tumor_grade == '2', 1,
                                             ifelse(is.na(sumfun_data_full_covariates$tumor_grade), NA,
                                                    0))
sumfun_data_full_covariates$grade3 <- ifelse(sumfun_data_full_covariates$tumor_grade == '3', 1,
                                             ifelse(is.na(sumfun_data_full_covariates$tumor_grade), NA,
                                                    0))
```

Calculating the p-value for the test of the spatial covariate:

```{r message=FALSE, warning=FALSE, error=FALSE, results='hide'}
CB_beta4 <- wildBS_CB(formula=outcome ~ patient_age + grade2 + grade3 + spatial_cov,
                      data=sumfun_data_full_covariates,
                      spatial_covars = c("spatial_cov"),
                      target = c("spatial_cov"),
                      form   = c("coef"),
                      covar_list = list(),
                      func = function(spatial_cov){
                        spatial_cov
                      },
                      B1=10,
                      B2=10,
                      alpha=.05,
                      re=NULL,
                      nthreads = 1,
                      id='patient_id')
```

```{r message=FALSE, warning=FALSE, error=FALSE}
M <- max(abs((CB_beta4$target_curve_estimates - 0)/CB_beta4$target_curve_SEs))
mean(CB_beta4$M>M) #p-value for the test of beta4(r)=0
```

## Melanoma Data

```{r }
data("melanoma_data", package='funboot')
```

```{r message=FALSE, warning=FALSE, error=FALSE, results='hide'}
set.seed(12345)
sumfun_data <- preprocess_data(data=melanoma_data,
                               from_cell="CD8- T cell",
                               to_cell="CD8+ T cell",
                               qc_cellcount_cutoff=20,
                               n_perm=5,
                               perm_yn=TRUE,
                               r_max=500,
                               inc=1,
                               image_dims=c(0,1200,0,1200),
                               summary_function='L',
                               verbose=TRUE)
sumfun_data$patient_id <- as.factor(sumfun_data$patient_id)
set.seed(12345)
#var_mat represents the values of the spatial covariate \hat{L}_{CD8-,Vas}(r)
var_mat_data <- preprocess_data(melanoma_data,
                                from_cell='CD8- T cell',
                                to_cell='Vasculature',
                                qc_cellcount_cutoff=20,
                                n_perm=5,
                                perm_yn=TRUE,
                                r_max=500,
                                inc=1,
                                image_dims=c(0,1200,0,1200),
                                summary_function='L',
                                verbose=TRUE)
var_mat_data$patient_id <- as.factor(var_mat_data$patient_id)
names(var_mat_data)[names(var_mat_data)=='L_obs'] <- 'var_mat'
var_mat_data$L_expect <- var_mat_data$L_pmean <- var_mat_data$outcome <- NULL
sumfun_data_full <- merge(sumfun_data, var_mat_data, by=c('image_number','r','patient_id','patient_age'),all=T)
sumfun_data_full$var_mat <- ifelse(is.na(sumfun_data_full$var_mat), 0, sumfun_data_full$var_mat)

covar_list <- list(Intercept = rep(1, 501),
                   var_mat = rep(0, 501))

CB_avascular <- wildBS_CB(formula=outcome ~ var_mat + s(patient_id, bs = "re"),
                         data=sumfun_data_full,
                         spatial_covars = c('var_mat'),
                         target = c("Intercept","var_mat"),
                         form   = c("coef","term"),
                         covar_list = covar_list,
                         func = function(Intercept, var_mat){
                           Intercept + var_mat
                         },
                         B1=10,
                         B2=10,
                         alpha=.05,
                         re='patient_id',
                         nthreads = 1,
                         id='patient_id')
```

```{r fig3, fig.cap="**Figure 2.** Spatial distributions of selected cell types in representative breast cancer images.", fig.width = 7, fig.height = 3, warning=FALSE, message=FALSE, results='hide'}
p <- funboot::plot_wildBS_CB(CB_avascular)

p +
  labs(
    title = "Avascular Regions",
    x = "r",
    y = "E[Y(r)|X]"
  ) +
  scale_color_manual(
    name = " ",
    values = c(
      "Estimated Target Curve" = "#F8766D",
      "Confidence Band" = "#00BA38"
    ),
    labels = c(
      "Estimated Target Curve" = "Estimated E[Y(r)|X]",
      "Confidence Band" = "95% Confidence Band"
    ),
    breaks = c("Estimated Target Curve", "Confidence Band")
  ) +
  coord_cartesian(ylim = c(-100, 200)) +
  theme(
    axis.title.y = element_text(size = 20),
    axis.title.x = element_text(size = 20),
    plot.title = element_text(size = 18),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 18),
    plot.subtitle = element_text(size = 18)
  )
```

## Conclusion
