---
title: "Analyzing multiplex imaging data using functional regression and wild bootstrap"
output: rmarkdown::html_vignette
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this tutorial, we demonstrate the utilities of our method, *funboot,* for analyzing cell type colocalization in multiplex imaging data.  We apply functions from our package to a data subset of imaging mass cytometry breast cancer data.

**Load package and dependencies**

```{r setup, error=FALSE, warning=FALSE, message=FALSE}
#devtools::install_github('carlyemiddleton/funboot')
library(funboot) 

#Dependencies for the analysis functions
library(spatstat.data)
library(spatstat.univar)
library(spatstat.geom)
library(spatstat.random)
library(spatstat.explore)
library(refund)
library(dplyr)
library(parallel)

#Dependencies for the plotting functions
library(ggplot2)
library(patchwork)

```

**Disclaimer:**  The purpose of this tutorial is to demonstrate how to use our functions.  We do not claim the results from this tutorial to be a full data analysis by which reliable biological conclusions should be drawn. Here we use a small number of bootstrap samples in order to save computation time for demonstration purposes, but our manuscript contains the full data analysis.    

## Breast Cancer Data

We analyze the 100-image subset of the popular Jackson-Fischer breast cancer dataset [@jackson-fischer], available in the R package [imcdatasets](https://bodenmillergroup.github.io/imcdatasets/).  The data subset with the relevant variables extracted is available within *funboot.* We load it below:  

```{r }
data("breastcancer_data_subset_100images")
```

We are interested in cell metaclusters 3 (T cells), 7 (Endothelial cells), 9 (Small circular stromal cells), and 16 (Proliferative epithelial cells).  Let's start by examining their distributions, using the **plot_images()** function from *funboot.*

```{r fig.width = 7, fig.height = 12}
pallete <- c('Other' = "gray", '7' = 'gold', '3' = "darkgreen", '9' = 'skyblue', '16'='cyan4')
plot.list <- funboot::plot_images(data = breastcancer_data_subset_100images, images.to.plot = c(13,181,246,362,
                                                                                       219,244,246,254) #the example images from the manuscript
                         , cell.types.to.plot=c(7,3,9,16), pallete=pallete)

(plot.list[[1]] + plot.list[[2]])/
(plot.list[[3]] + plot.list[[4]])/
(plot.list[[5]] + plot.list[[6]])/
(plot.list[[7]] + plot.list[[8]])

```

In the original publication, cell types 7 and 3 are observed to be colocalized.  We confirm this observation by fitting the concurrent functional regression model   

$$
\hat{L}_{iEndo,T}(r) - r = \beta_0(r) + \epsilon_i(r)
$$

and performing inference on the intercept $\beta_0(r)$.  The first step is to preprocess the raw data and calculate the appropriate spatial summary function, applying the necessary permutation adjustment and quality control cell count cutoff [@wilson; @spaceanova].  This can be carried out using our function **preprocess_data():**


```{r results='hide', eval=F}
sumfun.data <- funboot::preprocess_data(data=breastcancer_data_subset_100images, from.cell=7, to.cell=3,
                               qc.cellcount.cutoff=20, P=50, perm.yn=T,
                               R=500, inc=1, image.dims=c(0,1000,0,1000),
                               summary.function='L',seed=456)

```

We fit the model and examine the output:

```{r eval=F}
model <- funboot::fit_model(formula=outcome ~ 1, data=sumfun.data)
summary(model)

```

Both the constant intercept and smooth intercept from the **pffr()** output are positive; this is indicative of cell type interaction.  We calculate and plot the wild bootstrap confidence band of the sum of both intercepts using **lin_comb_CB()** and **plot_lin_comb_CB().**

```{r fig.width = 7, fig.height = 3, eval=F}
CBs <- funboot::lin_comb_CB(formula=outcome ~ 1, lin.comb=list(), 
                   data=sumfun.data, spatial.covars = NULL,
                   B1=10,B2=10,alpha=.05,re=NULL,seed=456,n_cores=NULL)
#Plot CBs
funboot::plot_lin_comb_CB(CBs)

```

Because the confidence band crosses zero, the conclusion from this demonstration would be that a significant amount of cell type interaction is present. (However, many more bootstrap samples are needed for accurate results.) We repeat the above for metaclusters 9 and 16, observing significant cell type avoidance:

```{r results='hide', eval=F}
sumfun.data.9.16 <- funboot::preprocess_data(data=breastcancer_data_subset_100images, from.cell=9, to.cell=16,
                               qc.cellcount.cutoff=20, P=50, perm.yn=T,
                               R=200, inc=1, image.dims=c(0,1000,0,1000),
                               summary.function='L',seed=456)
```

```{r fig.width = 7, fig.height = 3, eval=F}
model.9.16 <- funboot::fit_model(formula=outcome ~ 1, data=sumfun.data.9.16)
summary(model.9.16)

CBs.9.16 <- funboot::lin_comb_CB(formula=outcome ~ 1, lin.comb=list(), 
                   data=sumfun.data.9.16, spatial.covars = NULL,
                   B1=10,B2=10,alpha=.05,re=NULL,seed=456,n_cores=NULL)
#Plot CBs
funboot::plot_lin_comb_CB(CBs.9.16)

```

We next consider the model

$$
\widehat{L}_{i Endo, Pro}(r) - r =  \beta_0(r) + \beta_1(r)x_{iage} + \beta_2(r)x_{igrade2} + \beta_3(r)x_{igrade3} + \beta_4(r)\widehat{L}_{iEndo,TM}(r) + \epsilon_i(r)
$$
Below is the procedure for calculating the necessary summary functions for this model.  We first calculate  $\widehat{L}_{i Endo, Pro}(r)$ by specifying metacluster 7 as the ``from" cell type and metacluster 16 as the ``to" cell type.  We consider metaclusters 3,4,5, and 6 to be T cells or macrophages, and calculate $\widehat{L}_{iEndo,TM}(r)$.  We then need to merge the two summary function datasets together. 


```{r  , eval=F, results='hide'}
sumfun.data <- funboot::preprocess_data(data=breastcancer_data_subset_100images, from.cell=7, to.cell=16,
                                   qc.cellcount.cutoff=20, P=50, perm.yn=T,
                                   R=500, inc=1, image.dims=c(0,1000,0,1000), #use R=500 so that we can plot the L curves as they return to 0
                                   summary.function='L',seed=NULL)
control.cells <- c(3:6)
breastcancer_data_subset_100images$cell_type <- ifelse(breastcancer_data_subset_100images$cell_type
                                                  %in% control.cells, 3, breastcancer_data_subset_100images$cell_type) #combine the control cells into 1 type
spatialcov.data <- funboot::preprocess_data(data=breastcancer_data_subset_100images, from.cell=7, to.cell=3,
                               qc.cellcount.cutoff=20, P=50, perm.yn=T,
                               R=500, inc=1, image.dims=c(0,1000,0,1000),
                               summary.function='L',seed=NULL)
names(spatialcov.data)[names(spatialcov.data)=='L.obs'] <- 'spatial.cov'
spatialcov.data$L.expect <- spatialcov.data$L.pmean <- spatialcov.data$outcome <- NULL
sumfun.data.full <- merge(sumfun.data, spatialcov.data, by=c('image_number','r','patient_id','patient_age','tumor_grade'),all=T)
df <- data.frame(image_number = rep(unique(breastcancer_data_subset_100images$image_number), each=501),
                 r = rep(0:500, length(unique(breastcancer_data_subset_100images$image_number))) )
sumfun.data.full <- merge(df, sumfun.data.full, by=c('image_number','r'),all=T)
sumfun.data.full <- sumfun.data.full[!is.na(sumfun.data.full$outcome),] #Drop obs. with missing outcomes
sumfun.data.full$spatial.cov <- ifelse(is.na(sumfun.data.full$spatial.cov), 0, sumfun.data.full$spatial.cov) #K function = 0 when there are no j cells
```

Next, we are ready to calculate and plot wild bootstrap confidence bands for each coefficient function:

```{r  eval=F, fig.width = 7, fig.height = 3, results='hide'}
##Wild BS CB
sumfun.data.full$grade2 <- ifelse(sumfun.data.full$tumor_grade == '2', 1,
                       ifelse(is.na(sumfun.data.full$tumor_grade), NA, 0))
sumfun.data.full$grade3 <- ifelse(sumfun.data.full$tumor_grade == '3', 1,
                                  ifelse(is.na(sumfun.data.full$tumor_grade), NA, 0))
CBs <- funboot::wildBS_CB(formula=outcome ~ patient_age + grade2 + grade3 + spatial.cov,
                   data=sumfun.data.full,spatial.covars = 'spatial.cov',B=10,alpha=.05,seed=NULL)

funboot::plot_wildBS_CB(CBs)
```

Using **Ftest()**, we can jointly test the two indicator variables for tumor grade and return the p-value for the test.

```{r  eval=F, results='hide'}
##Wild BS Partial F test
out <- funboot::Ftest(formula.full = outcome ~ patient_age + grade2 + grade3 + spatial.cov, 
                      formula.red = outcome ~ patient_age + spatial.cov,
                      data = sumfun.data.full, spatial.covars = 'spatial.cov', B=10,seed=456)
```

```{r  eval=F}
out$p.value
```


## References









